---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial illustrates the process of estimating bias from control samples
using this estimate to calibrate the relative abundances of the control taxa in
samples whose true composition is unknown. The data we use is from the
synthetic-community colonization experiment of Leopold and Busby (2020).

> Leopold DR, Busby PE. 2020. Host Genotype and Colonist Arrival Order Jointly Govern Plant Microbiome Composition and Function. Curr Biol 30:3260-3266.e5. doi:10.1016/j.cub.2020.06.011

These authors performed a colonization experiment in which black cottonwood
(*Populus trichocarpa*) was inoculated with the fungal rust pathogen
*Melampsora × columbiana* and 8 species of foliar fungi. Fungal relative
abundances were measured using ITS amplicon sequencing. To enable
quantification of bias due to PCR bias and variation in ITS copy number, they
also created a set of DNA mocks of these fungi, which they measured along with
the experimental samples. We will use the mock samples to estimate the bias of
the 9 focal species, and then calibrate these species' abundances in the
full set of samples.

# Setup

First, let's set up our R environment. This tutorial uses phyloseq objects and
functions to store and manipulate the microbiome data, tidyverse packages for
data manipulation and plotting, and some add-ons to ggplot2 from the ggbeeswarm
and cowplot packages.
```{r setup}
# Tools for microbiome data
library(phyloseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
theme_set(theme_cowplot())

library(metacal); packageVersion("metacal")
```

Next, we download the source data from Leopold and Busby (2020), which is
available in [Zenodo record 3872145](https://zenodo.org/record/3872145). Set
`data_path` to wherever you would like the data to be downloaded.
```{r}
data_path <- here::here("vignettes", "data", "leopold2020host")
# To use a temporary directory:
# data_path <- file.path(tempdir(), "leopold2020host")
if (!dir.exists(data_path)) {
  dir.create(data_path, recursive = TRUE)
  download.file(
    "https://zenodo.org/record/3872145/files/dleopold/Populus_priorityEffects-v1.2.zip",
    file.path(data_path, "Populus_priorityEffects-v1.2.zip")
  )
  unzip(
    file.path(data_path, "Populus_priorityEffects-v1.2.zip"), 
    exdir = data_path
  )
}
```

The microbiome data is stored in a phyloseq object,
```{r}
ps <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/output/compiled/phy.rds") %>%
  readRDS %>%
  print
```
The (expected) actual compositions of the mock (control) samples can be
computed from a csv file that contains the dilution factors used during sample
construction. I will read in the csv file, convert it to a phyloseq object, and
then compute the relative abundances of each species from the reciprocals of
the dilution factors.
```{r}
mock_actual <- file.path(data_path, 
  "dleopold-Populus_priorityEffects-8594f7c/data/MockCommunities.csv") %>%
  read.csv(row.names = 1) %>%
  select(-Sym4) %>%
  as("matrix") %>%
  otu_table(taxa_are_rows = FALSE) %>%
  transform_sample_counts(function(x) close_elts(1 / x))
mock_taxa <- taxa_names(mock_actual)
```
Note, I have converted the actual relative abundances to proportions, but the
observed abundances in the `ps` object have the form of read counts,
```{r}
otu_table(ps) %>% prune_taxa(mock_taxa, .) %>% corner
```

Let's take a look at the sample data and taxonomy table in the phyloseq object,
```{r}
sam <- sample_data(ps) %>% as("data.frame") %>% as_tibble(rownames = "Sample")
tax <- tax_table(ps) %>% as("matrix") %>% as_tibble(rownames = "Taxon")
sam %>% glimpse
sam %>%
  count(Samp_type)
```
The control samples are the 10 samples with `Samp_type == "Mock"`.

Now let's look at the taxonomy table,
```{r}
tax %>% head
tax %>% filter(Taxon %in% mock_taxa) %>% select(Taxon, Family:Species)
ntaxa(ps) 
```
There are 219 OTUs, of which 9 have been assigned to the mock taxa (the
pathogen, here named "Melampsora", and the 8 foliar fungi). The names of the
mock taxa in `ps` have already been set to exactly match those in
`mock_actual`, which is a requirement for the steps that follow.

# Inspect mock community measurements

To get a sense of how the mock mixtures were created, we can take a look at the
expected proportions of the 9 taxa across the 10 samples,
```{r, fig.dim = c(7.5, 3)}
mock_actual %>%
  psmelt %>%
  mutate(
    across(OTU, factor, levels = mock_taxa),
    across(Sample, factor, levels = sample_names(mock_actual)),
  ) %>%
  ggplot(aes(Sample, OTU, fill = Abundance)) +
  geom_tile() +
  scale_fill_viridis_c(trans = "log10", breaks = c(0.02, 0.05, 1e-1, 0.2)) +
  theme_cowplot(font_size = 11)
```

All taxa are in every sample, but in varying proportions that range from around
0.01 to 0.3.

To visualize and estimate bias, we'll work with a phyloseq object of the
observed compositions subset to just the mock community samples and taxa,
```{r}
ps.mock <- ps %>% 
  subset_samples(Samp_type == "Mock") %>%
  prune_taxa(mock_taxa, .)
```
Next, let's compare the observed to actual taxon proportions in the mocks. I'll
first combine the observed and actual proportions in a data frame in "long"
format,
```{r}
props <- list(Actual = mock_actual, Observed = ps.mock) %>%
  map(transform_sample_counts, close_elts) %>%
  map_dfr(psmelt, .id = "Type") %>%
  select(Type, OTU, Sample, Abundance)
```
We can plot the observed vs. actual proportions with ggplot as follows,
```{r, fig.dim = c(6, 4)}
brks <- c(0, 1e-3, 1e-2, 1e-1, 0.3)
axes <- list(
  scale_y_continuous(trans = scales::pseudo_log_trans(1e-3), 
    breaks = brks, labels = brks),
  scale_x_continuous(trans = scales::pseudo_log_trans(1e-3),
    breaks = brks, labels = brks),
  coord_fixed()
)

props %>%
  pivot_wider(names_from = Type, values_from = Abundance) %>%
  ggplot(aes(Actual, Observed, color = OTU, shape = Sample == "Mock.5")) +
  geom_abline(color = "darkgrey") +
  geom_quasirandom() +
  axes +
  scale_shape_manual(values = c(16, 1)) +
  scale_color_brewer(type = "qual", palette = 3)
```

Some notes about this plot:

* Axes transformations: The multiplicative error created by bias can only be
  easily seen for low-proportion taxa if we use some form of log
  transformation. Options to consider include a log (or log10) transformation,
  logit transformation, and a pseudo-log transformation (which is linear at low
  values and logarithmic at higher values). Here I will use the pseudo-log
  transform since it can more easily handle any 0s that happen to be in the
  data than these other transforms. I set `1e-3` as the linear-to-log
  transition point; you should play around with this value with your own data.
  I saved the transformed axes in a variable `axes` to reuse in later plots.

* Because the actual proportions are clustered at just 4 values, I used
  `geom_quasirandom()` from the ggbeeswarm package to jitter the points.

* I have highlighted the sample Mock.5 as it was identified as an outlier by
  Leopold2020. It is the only sample where a taxon (Epicoccum) was undetected
  in a sample where it should have been present (the point at (0.01, 0)). Other
  than this non-detection, the observations of this sample seem in line with
  the others. I will include this sample in this tutorial, but we need to be
  aware that how we deal with this zero value may have undue influence on our
  bias estimate.

# Estimate bias

Metacal provides a high-level interface to estimating bias, `estimate_bias()`,
which expects two matrices or phyloseq objects with the observed and actual
abundances for a set of control samples. To obtain sensible estimates using
this function, we must first ensure that all taxa have an observed abundance
that is greater than 0 in samples where their actual abundance is greater than
0. We will therefore add a pseudocount of 1 to the observed counts to
remove 0s; other zero-replacement methods, such as `cmultRepl()` from
zCompositions, would also work. The estimation method used by `estimate_bias()`
also requires that taxa not supposed to be in a sample have an abundance of
exactly 0, but the `estimate_bias()` function will automatically zero-out these
observations for you.
```{r}
sum(otu_table(ps.mock) == 0)
ps.mock.pseudo <- ps.mock %>%
  transform_sample_counts(function(x) x + 1)
all(otu_table(ps.mock.pseudo) > 0)
```

Now we can estimate bias using `ps.mock.pseudo` as our observed abundances. We
use the `boot = TRUE` option to also generate bootstrap replicates that will
give us a way to quantify the uncertainty in our estimate.
```{r}
mc_fit <- estimate_bias(ps.mock.pseudo, mock_actual, boot = TRUE) %>% print
class(mc_fit)
```
`estimate_bias()` returns an "mc_bias_fit" object that can be interacted with
through commonly used S3 methods such as `print()`, `summary()`, and `coef()`.
The bias estimate can be extracted with `coef()`,
```{r}
bias <- coef(mc_fit) %>% print
```
The `summary()` method uses the bootstrap replicates in the "mc_bias_fit"
object to estimate the (geometric) standard error of the bias estimate,
```{r}
mc_fit.summary <- summary(mc_fit)
print(mc_fit.summary)
```
producing a data frame with the estimate and standard error that can be
accessed with
```{r}
coef_tb <- mc_fit.summary$coefficients
```

Let's use this data frame to plot the bias estimate with two geometric standard
errors,

```{r, fig.dim = c(5,3)}
coef_tb %>%
  mutate(taxon = fct_reorder(taxon, estimate)) %>%
  ggplot(aes(taxon, estimate, 
      ymin = estimate / gm_se^2, ymax = estimate * gm_se^2)) +
  geom_hline(yintercept = 1, color = "grey") +
  geom_pointrange() +
  scale_y_log10() +
  coord_flip()
```

This plot shows that these estimates are consistent but not identical with
those in Leopold2020 Fig S2. The difference is due to our inclusion of the
Mock.5 sample, which leads to a lower point estimate and larger standard error
for *Epicoccum*.

## Check model fit

The fitted values of the model can be accessed with `fitted()`; these are the
proportions that we predict we'd observe for the control samples given the
actual proportions and the estimated bias. The `fitted()` function returns a
matrix with samples as rows and taxa as columns, which I'll convert into a
phyloseq object.
```{r}
observed.fitted <- fitted(mc_fit) %>% otu_table(taxa_are_rows = FALSE)
```
To plot the fitted values against the observed values, I'll add the fitted
proportions to our earlier `props` data frame,
```{r}
props.fitted <- bind_rows(
  props, 
  psmelt(observed.fitted) %>% add_column(Type = "Fitted")
)
```
Next, I'll use some tidyr manipulations and ggplot to plot the observed
proportions vs. the actual and fitted predictions,
```{r, fig.dim = c(8.5, 4)}
props.fitted %>%
  pivot_wider(names_from = Type, values_from = Abundance) %>%
  pivot_longer(c(Fitted, Actual), names_to = "Type", values_to = "Predicted") %>%
  ggplot(aes(Predicted, Observed, color = OTU)) +
  geom_abline(color = "darkgrey") +
  geom_point() +
  axes + 
  facet_wrap(~Type) +
  scale_color_brewer(type = "qual", palette = 3)
```

This plot shows that the fitted model is doing a good job of explaining the
error across the full range of predicted proportions, though it is slightly
underpredicting the observed proportions at low values.

# Calibrate

Now we can use the estimated bias to calibrate the relative abundances of
the mock taxa in all samples. Metacal provides the `calibrate()` function for
this purpose, which takes as its arguments the observed compositions and the
bias estimate returned by `coef(mc_fit)`. 

There are a few things to keep in mind about `calibrate()`.

* Because of the relative nature of microbiome measurement and the metacal bias
  estimate, it is only possible to calibrate the relative abundances of the
  taxa whose bias has been estimated. For this reason, `calibrate()`
  automatically drops taxa that are missing from the bias estimate.

* A deterministic calibration method is used in which the observed abundances
  are simply divided by the bias and (optionally) normalized to sum to 1.
  Uncertainty in the bias estimate and in the observed abundances due to the
  noise inherent in microbiome measurement (such as random sampling of
  sequencing reads) is not accounted for.

* An implication is that taxa whose observed abundance is 0 will have a
  calibrated abundance of 0. Whether and how you replace 0s prior to
  calibration may have a significant impact on downstream analyses.

Here I will use a pseudocount to make all the taxa abundances positive prior to
calibration.
```{r}
ps.pseudo <- transform_sample_counts(ps, function(x) x + 1)
ps.pseudo.cal <- calibrate(ps.pseudo, bias) %>% print
```

To verify that calibration worked as expected, let's compare the observed,
calibrated, and actual taxon proportions in the 10 control samples. Again, I'll
start by adding the calibrated proportions to the data frame with observed and
actual proportions,

```{r}
ps.pseudo.cal.mock <- ps.pseudo.cal %>% 
  subset_samples(Samp_type == "Mock")
props.cal <- bind_rows(
  props, 
  psmelt(ps.pseudo.cal.mock) %>% add_column(Type = "Calibrated")
) %>%
  select(Type, OTU, Sample, Abundance)
```

First, let's look at bar plots comparing the observed, calibrated, and actual
compositions,

```{r, fig.dim = c(8, 6)}
props.cal %>%
  mutate(
    across(Type, fct_rev),
    across(Sample, factor, levels = sample_names(mock_actual)),
    ) %>%
  rename(Proportion = Abundance) %>%
  ggplot(aes(Type, Proportion, fill = OTU)) +
  geom_col(width = 0.7) +
  scale_fill_brewer(type = "qual", palette = 3) +
  facet_wrap(~Sample, ncol = 5) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  )
```

We can see that the calibrated compositions are typically substantially closer
to the actual proportions than the original observations. We can also plot the
calibrated vs. actual (pseudo-log-transformed) proportions and compare to the
earlier plot of observed vs. actual.

```{r, fig.dim = c(7, 4)}
props.cal %>%
  pivot_wider(names_from = Type, values_from = Abundance) %>%
  pivot_longer(c(Observed, Calibrated), 
    names_to = "Type", values_to = "Proportion") %>%
  mutate(across(Type, fct_relevel, "Observed")) %>%
  ggplot(aes(Actual, Proportion, color = OTU)) +
  geom_abline(color = "darkgrey") +
  geom_quasirandom() +
  facet_wrap(~Type) +
  axes +
  scale_color_brewer(type = "qual", palette = 3)
```

This plot has essentially same information as the earlier plot examining the
model fit, but seen from a different angle. We can see that calibration in this
dataset clearly reduces the relative error in proportions across the tested
range of actual proportions, but leads to estimates that are systematically too
low at low proportions.

# Session info

```{r}
sessioninfo::session_info()
```
